Structured Versus Decorated Cospans

### 2020-12-30 cospans.tex
Update cospans.tex

\begin{abstract}
One goal of applied category theory is to understand open systems.  We compare two ways of describing open systems as cospans equipped with extra data.    First, given a functor $L \maps \A \to \X$, a `structured cospan' is a diagram in $\A$ of the form $L(a) \rightarrow x \leftarrow L(b)$.  If $\A$ and $\X$ have finite colimits and $L$ preserves them, it is known that there is a symmetric monoidal double category whose objects are those of $\A$ and whose horizontal 1-cells are structured cospans.   Second, if $\A$ has finite colimits and $F \maps (\A,+) \to (\Cat,\times)$ is a symmetric lax monoidal pseudofunctor , a `decorated cospan' is a diagram in $\A$ of the form $a \rightarrow s \leftarrow b$ together with an object of $F(s)$.   Generalizing the work of Fong, we show that in this situation there is a symmetric monoidal double category whose objects are those of $\A$ and whose horizontal 1-cells are decorated cospans.   We prove that under certain conditions these two constructions become isomorphic when we take $\X = \int F$ to be the Grothendieck category of $F$.  We illustrate these ideas with applications to electrical circuits, Petri nets, dynamical systems and epidemiological modeling.
\end{abstract}

@@

Here we give sufficient conditions for a decorated cospan double category to be equivalent---and in fact, isomorphic---to a structured cospan double category.  Suppose $\A$ has finite colimits and $F \maps (\A , +) \to (\Cat, \times)$ is a symmetric lax monoidal pseudofunctor.   Any such pseudofunctor gives an opfibration $U \maps \X \to \A$ where $\X = \inta F$ is defined by the Grothendieck construction.  Let $\Rex$ be the 2-category of categories with finite colimits, functors preserving finite colimits, and natural transformations.  We show that if $F$ factors through $\mathbf{Rex}$ as a pseudofunctor, the opfibration $U \maps \X \to \A$ is also a right adjoint.  From the accompanying left adjoint $L \maps \A \to \X$, we  construct a symmetric monoidal double category ${}_L \lCsp(\X)$ of structured cospans.  Finally, in \cref{thm:equiv} we prove that this structured cospan double category ${}_L \lCsp(\X)$ is isomorphic to the decorated cospan double category $F \lCsp$.

@@

In \cref{DecCospansDoubleCat}, we construct the symmetric monoidal double category $F\lCsp$ and we define maps between decorated cospan double categories. In \cref{EquivDoubleCats}, we briefly review the structured cospans framework and prove that the double categorical versions of decorated cospans and structured cospans are isomorphic under suitable conditions. In \cref{spinoffs}, we establish the isomorphism between structured and decorated cospans at the level of bicategories and categories (via decategorification).
Finally, in \cref{Applications}, we describe applications to electrical circuits, Petri nets, epidemiology and dynamical systems.

@@

The first part formulates the existence of colimits \emph{locally} in each fiber, and if we let $\J$ range over all finite categories it says that the corresponding pseudofunctor $F \maps \A \to \Cat$ lands in the sub-2-category $\Rex$. The second part formulates the existence of colimits \emph{globally} in the total category $\inta F$, and if we let $\J$ range over all finite categories it says that $\X$ has finite colimits and $U$ preserves all finite colimits. As an example, suppose that $\A$ has pushouts and a Grothendieck opfibration $U \maps \inta F \to\A$ has fiberwise pushouts preserved by its reindexing functors. We can construct pushouts in $\inta F$ as follows: 
\begin{equation}\label{eq:globalpushout}
 \begin{tikzcd}[ampersand replacement=\&,column sep=1in,row sep=.5in]
\& (a+_bc,w)\ar[dd,phantom, very near start,"\upback"description] \& \\
(a,x)\ar[ur] \&\& (c,z)\ar[ul] \\
\& (b,y)\ar[ul,"{\begin{cases}f\maps b\to a &\textrm{in }\A \\k\maps F(f)y\to x &\textrm{in }F(a)\end{cases}}"description]\ar[ur,"{\begin{cases}g\maps b\to c &\textrm{in }\A \\\ell\maps F(g)y\to z &\textrm{in }F(c)\end{cases}}"description] \&
 \end{tikzcd}
\end{equation}
where $a+_b c$ is defined using the pushout in $\A$ shown at left below, and $w$ is formed as the pushout in the fiber $F(a+_bc)$ at the right below:
\begin{displaymath}
\begin{tikzcd}
& a+_bc\ar[dd,phantom, very near start,"\upback"description] & \\
a\ar[ur,"v_a"] && c\ar[ul,"v_c"'] \\
& b\ar[ul,"f"]\ar[ur,"g"'] &
\end{tikzcd}\qquad
\begin{tikzcd}
 & w \ar[dd,phantom, very near start,"\upback"description] & \\
 F(v_a)x\ar[ur] & & F(v_c)z\ar[ul] \\
 & F(v_a)(F(f)(y))\cong F(v_c)(F(g)(y))\ar[ul,"F(v_a)k"]\ar[ur,"F(v_c)\ell"'] &
 \end{tikzcd}
\end{displaymath}
In a similar way we can construct the initial object and coproducts in $\int F$ from those in the fibers, namely
\begin{equation}\label{eq:fibrewisecoprod}
(a,x)+(b,y)=\left(a+ b,F(\iota_a)x+F(\iota_b)y\right), \qquad 
 0_{\inta F}=\left(0_A,0_{F(0_A)}\right)
\end{equation}
where $\iota_a \maps a\to a+b$ and $\iota_b \maps a \to a+b$ are the sum inclusions in $\A$ and the sum on the second variable is in the fiber $F(a+b)$.

@@

In \cref{thm:equiv} we gave conditions under which the symmetric monoidal double category of \emph{decorated} cospans $F\lCsp$ is isomorphic to the  symmetric monoidal double category of \emph{structured} cospans ${}_L \lCsp(\int F)$.   We now show that under the same conditions we get an isomorphism of symmetric monoidal bicategories, and of categories.

>>>>

\begin{abstract}
One goal of applied category theory is to understand open systems.  We compare two ways of describing open systems as cospans equipped with extra data.    First, given a functor $L \maps \A \to \X$, a `structured cospan' is a diagram in $\A$ of the form $L(a) \rightarrow x \leftarrow L(b)$.  If $\A$ and $\X$ have finite colimits and $L$ preserves them, it is known that there is a symmetric monoidal double category whose objects are those of $\A$ and whose horizontal 1-cells are structured cospans.   Second, if $\A$ has finite colimits and $F \maps (\A,+) \to (\Cat,\times)$ is a symmetric lax monoidal pseudofunctor , a `decorated cospan' is a diagram in $\A$ of the form $a \rightarrow s \leftarrow b$ together with an object of $F(s)$.   Generalizing the work of Fong, we show that in this situation there is a symmetric monoidal double category whose objects are those of $\A$ and whose horizontal 1-cells are decorated cospans.   We prove that under certain conditions these two constructions become isomorphic when we take $\X = \inta F$ to be the Grothendieck category of $F$.  We illustrate these ideas with applications to electrical circuits, Petri nets, dynamical systems and epidemiological modeling.
\end{abstract}

@@

Here we give sufficient conditions for a decorated cospan double category to be equivalent---and in fact, isomorphic---to a structured cospan double category.  Suppose $\A$ has finite colimits and $F \maps (\A , +) \to (\Cat, \times)$ is a symmetric lax monoidal pseudofunctor.   Any such pseudofunctor gives an opfibration $U \maps \X \to \A$ where $\X = \inta F$ is defined by the Grothendieck construction.  Let $\Rex$ be the 2-category of categories with finite colimits, functors preserving finite colimits, and natural transformations.  We show that if $F$ factors through $\Rex$ as a pseudofunctor, the opfibration $U \maps \X \to \A$ is also a right adjoint.  From the accompanying left adjoint $L \maps \A \to \X$, we  construct a symmetric monoidal double category ${}_L \lCsp(\X)$ of structured cospans.  Finally, in \cref{thm:equiv} we prove that this structured cospan double category ${}_L \lCsp(\X)$ is isomorphic to the decorated cospan double category $F \lCsp$.   In fact, they are isomorphic as symmetric monoidal double categories.

@@

In \cref{DecCospansDoubleCat}, we construct the symmetric monoidal double category $F\lCsp$ and we define maps between decorated cospan double categories. In \cref{EquivDoubleCats}, we briefly review the structured cospans framework and prove that the double categorical versions of decorated cospans and structured cospans are isomorphic under suitable conditions. In \cref{spinoffs}, we establish the isomorphism between structured and decorated cospans at the level of bicategories and categories (via decategorification).
Finally, in \cref{Applications}, we describe applications to electrical circuits, Petri nets, dynamical systems and epidemiological modeling.

@@

The first part formulates the existence of colimits \emph{locally} in each fiber, and if we let $\J$ range over all finite categories it says that the corresponding pseudofunctor $F \maps \A \to \Cat$ lands in the sub-2-category $\Rex$. The second part formulates the existence of colimits \emph{globally} in the total category $\inta F$, and if we let $\J$ range over all finite categories it says that $\X$ has finite colimits and $U$ preserves all finite colimits. As an example, suppose that $\A$ has pushouts and a Grothendieck opfibration $U \maps \inta F \to\A$ has fiberwise pushouts preserved by its reindexing functors. We can construct pushouts in $\inta F$ as follows: 
\begin{equation}\label{eq:globalpushout}
 \begin{tikzcd}[ampersand replacement=\&,column sep=1in,row sep=.5in]
\& (a+_bc,w)\ar[dd,phantom, very near start,"\upback"description] \& \\
(a,x)\ar[ur] \&\& (c,z)\ar[ul] \\
\& (b,y)\ar[ul,"{\begin{cases}f\maps b\to a &\textrm{in }\A \\k\maps F(f)y\to x &\textrm{in }F(a)\end{cases}}"description]\ar[ur,"{\begin{cases}g\maps b\to c &\textrm{in }\A \\\ell\maps F(g)y\to z &\textrm{in }F(c)\end{cases}}"description] \&
 \end{tikzcd}
\end{equation}
where $a+_b c$ is defined using the pushout in $\A$ shown at left below, and $w$ is formed as the pushout in the fiber $F(a+_bc)$ at the right below:
\begin{displaymath}
\begin{tikzcd}
& a+_bc\ar[dd,phantom, very near start,"\upback"description] & \\
a\ar[ur,"v_a"] && c\ar[ul,"v_c"'] \\
& b\ar[ul,"f"]\ar[ur,"g"'] &
\end{tikzcd}\qquad
\begin{tikzcd}
 & w \ar[dd,phantom, very near start,"\upback"description] & \\
 F(v_a)x\ar[ur] & & F(v_c)z\ar[ul] \\
 & F(v_a)(F(f)(y))\cong F(v_c)(F(g)(y))\ar[ul,"F(v_a)k"]\ar[ur,"F(v_c)\ell"'] &
 \end{tikzcd}
\end{displaymath}
In a similar way we can construct the initial object and coproducts in $\inta F$ from those in the fibers, namely
\begin{equation}\label{eq:fibrewisecoprod}
(a,x)+(b,y)=\left(a+ b,F(\iota_a)x+F(\iota_b)y\right), \qquad 
 0_{\inta F}=\left(0_A,0_{F(0_A)}\right)
\end{equation}
where $\iota_a \maps a\to a+b$ and $\iota_b \maps a \to a+b$ are the sum inclusions in $\A$ and the sum on the second variable is in the fiber $F(a+b)$.

@@

In \cref{thm:equiv} we gave conditions under which the symmetric monoidal double category of \emph{decorated} cospans $F\lCsp$ is isomorphic to the  symmetric monoidal double category of \emph{structured} cospans ${}_L \lCsp(\inta F)$.   We now show that under the same conditions we get an isomorphism of symmetric monoidal bicategories, and of categories.
